 

### 2023-04-26 markdown/methods.Rmd
first draft flowchart for analysis pipeline

## Data
```{r Dataflow,  echo=FALSE}
library(DiagrammeR)
library(webshot)

DiagrammeR::grViz("digraph {
	graph [layout  =  dot,
	       rankdir  = TB]

node [shape = rectangle]
rawData [label = 'msData.raw']
node [shape = oval]
MQ [label = 'MaxQuant']
node [shape = invhouse]
MQ_fd [label = 'feature detection']
node [shape = cylinder]
MQ_ds [label = 'Database search']
node [shape = invhouse]
MQ_fdr [label = 'FDR calculation']
MQ_quant [label = 'Protein Quantification']
node [shape = rectangle]
MQoutput [label = 'evidence.txt']
sampleAnnotate [label = 'sample_annotation.txt']
node [shape = oval]
ProteomicsWorkbench [label = 'Proteomics Workbench']

rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
sampleAnnotate -> ProteomicsWorkbench
}",
height = 500)
```

@@

### MS-Spectrum

@@

### Interpretation of the data

#### Data acquisition

@@

##### Three-dimensional peak detection

@@

##### Deisotoping

@@

##### Label detection

@@

##### Improving peptide mass accuracy

@@

##### Peptide search

@@

##### Protein assembly

@@

#### Data processing

Further analysis can be done with R and respective packages such as bioconductor. Since there is no state of the art established, analysis varies upon experimental design.

##### Reading the data

@@

##### Object oriented programming

@@

##### Zero values
Peptides with low abundance are often set to zero during analysis. However, assigning a value of zero may incorrectly suggest that the sample does not contain the respective peptide. Given that it is highly unlikely for a biological cell of a comparable type and function to not contain a particular protein, replacing the zero value with "not applicable" (NA) is crucial for understanding and interpreting MS data.

##### Exclude reverse matches/contaminants
Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.

##### Filter according to precursor ion fraction (PIF)
During mass spectrometry, the ions detected in MS1 are further fragmented through collision during multiple MS runs. The resulting product ions are derived from precursor ions (also known as mother ions or parental ions). Contaminant peptides can co-migrate in this process and can be distinguished by the lower fraction of their respective precursor ions \citep{Tannous2020}. These peptides need to be filtered out during the analysis pipeline. A cutoff value, referenced in the SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface, but it can be adjusted according to the needs of the biologist.

##### Filter by q-value
The next step for quality control is the exclusion of samples with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score. The peptides spectrum matching score is defined as -10log10(p). Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda , an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

##### Peptide spectrum match (PSM) aggregation to peptides 
In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function. In the context of processing from peptides to spectrum matches, the desired column is the peptide sequence. To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.

##### Join assays when observing multiple comparable batches at once
Sample size is often a limiting factor in hypothesis testing. A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to an early end of analyses. To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.

##### Calculate reporter ion intensity (RI) and filter according to median RI
Columns which do not meet the desired intensity can be filtered by a threshold set on the RI. The median RI can also be used to check if an entire channel has a lower detection level. This can be due two reasons. One is the expression level of the given proteins in a cell. Meaning, that the expression of the observed cell type is simply lower than the other type. Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.

##### Calculate and filter according to median coefficient of variation (CV) per cell/channel
Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data. The coefficient of variation of a peptide is considered as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins). Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.

##### Remove peptides with high missing rate
Although missing value imputation can be performed during the analysis of multiple batches, peptides with missing detections across channels can be problematic for quantification. The proteomic composition of a biological sample is similar between replicates and even across groups. However, the threshold of missingness (described as a fraction of the row) can be set in the user interface and adjusted to enable different experimental designs.

##### Aggregation of peptides to proteins 
Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.

##### Transformation of protein expression data
Data transformation applies a function to each value of a matrix or array, so that: 
$$
y_i = f(x_i)
$$
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill dependencies for statistical testing. In the shiny application various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et. Al mentioned in their SCoPE2 publication citep\{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and the remove skewness in the dataset. This transformation was used as a reference when comparing methods. However using the logarithm to the base 10 can be easy interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method which is also implemented in the application. The boxcox transformation developed 1964 by Box and Cox citep\{Sakia1992} applies a recursion on the data by determining the parameter lambda. Depending on the size of lambda a certain function will be applied on each value of the expression set in order to introduce normality. As already mentioned transformations applied by the BoxCox method range from log10 to taking the square root of each value

>>>>

## Data
```{r Dataflow, eval=FALSE, include=FALSE}
# library(DiagrammeR)
# library(webshot)
# 
# DiagrammeR::grViz("digraph {
# 	graph [layout  =  dot,
# 	       rankdir  = TB]
# 
# node [shape = rectangle]
# rawData [label = 'msData.raw']
# node [shape = oval]
# MQ [label = 'MaxQuant']
# node [shape = invhouse]
# MQ_fd [label = 'feature detection']
# node [shape = cylinder]
# MQ_ds [label = 'Database search']
# node [shape = invhouse]
# MQ_fdr [label = 'FDR calculation']
# MQ_quant [label = 'Protein Quantification']
# node [shape = rectangle]
# MQoutput [label = 'evidence.txt']
# sampleAnnotate [label = 'sample_annotation.txt']
# node [shape = oval]
# ProteomicsWorkbench [label = 'Proteomics Workbench']
# 
# rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
# sampleAnnotate -> ProteomicsWorkbench
# }",
# height = 500)
```

@@

#### MS-Spectrum

@@

@@

#### Three-dimensional peak detection

@@

#### Deisotoping

@@

#### Label detection

@@

#### Improving peptide mass accuracy

@@

#### Peptide search

@@

#### Protein assembly

@@

### Data processing

Further analysis is done with R and respective packages such as bioconductor. Since there is no state of the art established, analysis varies upon experimental design. The workflow of the analysis will be processed in a so called pipeline streamlining the data through steps where individual results can be observed in visualizations and indivdual calculations will be adapted according to user demands and experimental properties.

```{r data_processing_pipeline_flowchart, echo=FALSE, fig.height=10}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

opts_box_setting <- gpar(fill="lightyellow", col="darkblue")
opts_txt_setting <- gpar(col="darkblue", cex=1)

# input files
evidence_file <- boxGrob({glue("evidence file",
                           .sep = "\n")})
sample_annotation_file <- boxGrob({glue("sample annotation",
                           .sep = "\n")})

read_data <- boxGrob({glue("read data into q-feature object",
                          .sep = "\n")})

quality_control_1 <- boxGrob({glue("quality control 1",
                          .sep = "\n")})
qc1_content <- boxGrob("filter out pool samples\n
                             replace zeros with NA\n
                             filter out contaminants\n
                             filter by q-value cutoff\n",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

aggregate_psms <-  boxGrob({glue("aggregate peptide spectrum matches to
                                 peptides\n
                                 join multiple batches",
                          .sep = "\n")})

quality_control_2 <- boxGrob({glue("quality control 2",
                          .sep = "\n")})
qc2_content <- boxGrob("filter by reporter ion intensity cutoff\n
                             filter by pepitde covariance to razor protein\n
                             remove peptides by missing rate\n",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

aggregate_peps <- boxGrob({glue("aggregate peptides to proteins",
                          .sep = "\n")})

transform_data <- boxGrob({glue("transform data",
                          .sep = "\n")})
transform_opts <- boxGrob("options:\n
                                log2\n
                                log10\n
                                sqrt\n
                                quadratic\n
                                BoxCox\n
                                None",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

normalize_data <- boxGrob({glue("normalize data",
                          .sep = "\n")})
normalize_opts <- boxGrob("options:\n
                                SCoPE2\n
                                CONSTANd\n
                                None",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

missing_value_handling <- boxGrob({glue("missing value handling",
                          .sep = "\n")})
missing_value_opts <- boxGrob("options:\n
                                    KNN\n
                                    drop rows\n
                                    replace with mean\n
                                    replace with median\n
                                    replace with 0",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

batch_correction <- boxGrob({glue("batch correction",
                          .sep = "\n")})
batch_correction_opts <- boxGrob("options:\n
                                       ComBat",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

dim_red <- boxGrob({glue("dimensionality reduction",
                          .sep = "\n")})
dim_red_actions <- boxGrob("procedures:\n
                                 PCA\n
                                 UMAP",
                       txt_gp = opts_box_setting,
                       box_gp = opts_box_setting,
                       just="left"
                       )

stat_module <- boxGrob({glue("statistics module",
                          .sep = "\n")})

grid.newpage()
vert <- {spreadVertical(files = evidence_file,
                       read_data = read_data,
                       quality_control_1 = quality_control_1,
                       aggregate_psms,
                       quality_control_2 = quality_control_2,
                       aggregate_peps,
                       transform_data = transform_data,
                       normalize_data = normalize_data,
                       missing_value_handling = missing_value_handling,
                       batch_correction = batch_correction,
                       dim_red = dim_red,
                       stat_module
)}
input <- alignVertical(reference = vert$files,
                      evidence_file, sample_annotation_file) %>%
  spreadHorizontal()
vert$files <- NULL

# right side
qc1_content <- moveBox(qc1_content,
                    x = .8,
                    y = coords(vert$quality_control_1)$top)

transform_opts <- moveBox(transform_opts,
                    x = .9,
                    y = coords(vert$transform_data)$top)

missing_value_opts <- moveBox(missing_value_opts,
                    x = .7,
                    y = coords(vert$missing_value_handling)$top)

dim_red_actions <- moveBox(dim_red_actions,
                    x = .9,
                    y = coords(vert$dim_red)$top)

# left side
qc2_content <- moveBox(qc2_content,
                    x = .2,
                    y = coords(vert$quality_control_2)$top)

normalize_opts <- moveBox(normalize_opts,
                    x = .2,
                    y = coords(vert$normalize_data)$top)

batch_correction_opts <- moveBox(batch_correction_opts,
                    x = .2,
                    y = coords(vert$batch_correction)$top)

for (i in 1:(length(vert) - 1)) {
  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
    print
}

# connect boxes
connectGrob(input[[1]], vert$read_data, type = "N")
connectGrob(input[[2]], vert$read_data, type = "N")

connectGrob(vert$quality_control_1, qc1_content, type = "horizontal", arrow_obj = arrow(angle=0))

connectGrob(vert$quality_control_2, qc2_content, type = "horizontal", arrow_obj = arrow(angle=0))

connectGrob(vert$transform_data, transform_opts, type = "horizontal", arrow_obj = arrow(angle=0))

connectGrob(vert$normalize_data, normalize_opts, type = "horizontal", arrow_obj = arrow(angle=0))

connectGrob(vert$missing_value_handling, missing_value_opts, type = "horizontal", arrow_obj = arrow(angle=0))

connectGrob(vert$batch_correction, batch_correction_opts, type = "horizontal", arrow_obj = arrow(angle=0))

connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))

# Print boxes
input
vert
qc1_content
qc2_content
transform_opts
normalize_opts
missing_value_opts
batch_correction_opts
dim_red_actions
```

#### Reading the data

@@

#### Object oriented programming

@@

#### Zero values
Peptides with low abundance are often set to zero during analysis. However, assigning a value of zero may incorrectly suggest that the sample does not contain the respective peptide. Given that it is highly unlikely for a biological cell of a comparable type and function to not contain a particular protein, replacing the zero value with "not applicable" (NA) is crucial for understanding and interpreting MS data.

#### Exclude reverse matches/contaminants
Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.

#### Filter according to precursor ion fraction (PIF)
During mass spectrometry, the ions detected in MS1 are further fragmented through collision during multiple MS runs. The resulting product ions are derived from precursor ions (also known as mother ions or parental ions). Contaminant peptides can co-migrate in this process and can be distinguished by the lower fraction of their respective precursor ions \citep{Tannous2020}. These peptides need to be filtered out during the analysis pipeline. A cutoff value, referenced in the SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface, but it can be adjusted according to the needs of the biologist.

#### Filter by q-value
The next step for quality control is the exclusion of samples with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score. The peptides spectrum matching score is defined as -10log10(p). Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda , an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

#### Peptide spectrum match (PSM) aggregation to peptides 
In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function. In the context of processing from peptides to spectrum matches, the desired column is the peptide sequence. To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.

#### Join assays when observing multiple comparable batches at once
Sample size is often a limiting factor in hypothesis testing. A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to an early end of analyses. To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.

#### Calculate reporter ion intensity (RI) and filter according to median RI
Columns which do not meet the desired intensity can be filtered by a threshold set on the RI. The median RI can also be used to check if an entire channel has a lower detection level. This can be due two reasons. One is the expression level of the given proteins in a cell. Meaning, that the expression of the observed cell type is simply lower than the other type. Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.

#### Calculate and filter according to median coefficient of variation (CV) per cell/channel
Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data. The coefficient of variation of a peptide is considered as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins). Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.

#### Remove peptides with high missing rate
Although missing value imputation can be performed during the analysis of multiple batches, peptides with missing detections across channels can be problematic for quantification. The proteomic composition of a biological sample is similar between replicates and even across groups. However, the threshold of missingness (described as a fraction of the row) can be set in the user interface and adjusted to enable different experimental designs.

#### Aggregation of peptides to proteins 
Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.

#### Transformation of protein expression data
Data transformation applies a function to each value of a matrix or array, so that: 
$$
y_i = f(x_i)
$$
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill dependencies for statistical testing. In the shiny application various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et. Al mentioned in their SCoPE2 publication citep\{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and the remove skewness in the dataset. This transformation was used as a reference when comparing methods. However using the logarithm to the base 10 can be easy interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method which is also implemented in the application. The boxcox transformation developed 1964 by Box and Cox citep\{Sakia1992} applies a recursion on the data by determining the parameter lambda. Depending on the size of lambda a certain function will be applied on each value of the expression set in order to introduce normality. As already mentioned transformations applied by the BoxCox method range from log10 to taking the square root of each value
